### 2020-07-30 latex/symbols.tex
Added Q to the symbols
Major edits to VNHC
Added several new VNHC-related references, updated the whole chapter of
VNHCs, and implemented my new notation using the matrix kronecker
product. Also added further justfication as to why we have our own
definition of VNHCs, which needs more work. Next, we need to update the
proof on why a VNHC is regular iff it is h(q,pu) and satisfies that
matrix rank definition.

% Define a table for the symbols using the booktabs convention
\noindent
\begin{table}[h!]
\resizebox{\textwidth}{!}{%
\centering
\begin{tabular}{ll}
   \toprule
   Symbol & Definition \\
   \midrule
   \(\n\) & The index set \(\{1,\ldots,n\}\) of natural numbers up to \(n\). \\
   \(\R^n\) & Real numbers in \(n\) dimensions. \\
   \(\Rt{T}\) & Real numbers modulo \(T > 0\), with \(\Rt{\infty} = \R\). \\
   \(\mathbb{S}^1\) & The unit circle, equivalent to \(\Rt{2\pi}\). \\
   \(\R^{n \times m}\) & The space of real-valued matrices with \(n\) rows and \(m\) columns. \\
   \(\Id{n}\) & The \(n \times n\) identity matrix. \\
   \(\Zmat{n \times m}\) & The \(n \times m\) matrix of all zeros. \\
   \(M_i\) & If \(M\) is a vector, the \(i\)th element of \(M\). \\
         & If \(M\) is a matrix, the \(i\)th column of \(M\). \\
   \(M_{i,j}\) & The value of row \(i\), column \(j\) for the matrix \(M\). \\
   \(\dot{x}\) & Derivative of \(x\) with respect to time \(t\). \\
   \(\nabla_v F\) & Gradient of the \(\R\)-valued function \(F\) with respect to \(v\). \\
   \(dF_v\) & Total differential (Jacobian) of the function \(F\) with respect to \(v\).\\
           & If \(F\) is \(\R\)-valued, equivalent to \((\nabla_v F)\tpose\). \\
   \(\Hess F\) & Hessian matrix of double derivatives of \(F : \R^n \rightarrow \R\). \\
   \(\partial_v \partial_w F\) & Derivative matrix of \(F : \R^n \times \R^m \rightarrow \R\), with \((i,j)\) element \(\ppdiff{F}{v_i}{w_j}\). \\
   \(\delta_{i,j}\) & The Kronecker delta: 1 if \(i = j\) and 0 otherwise. \\
   \bottomrule
\end{tabular}

>>>>

% Define a table for the symbols using the booktabs convention
\noindent
\begin{table}[h!]
\resizebox{\textwidth}{!}{%
\centering
\begin{tabular}{ll}
   \toprule
   Symbol & Definition \\
   \midrule
   \(\n\) & The index set \(\{1,\ldots,n\}\) of natural numbers up to \(n\). \\
   \(\R^n\) & Real numbers in \(n\) dimensions. \\
   \(\Rt{T}\) & Real numbers modulo \(T > 0\), with \(\Rt{\infty} = \R\). \\
   \(\mathbb{S}^1\) & The unit circle, equivalent to \(\Rt{2\pi}\). \\
   \(\mathcal{Q}\) & The configuration manifold of a system. \\
   \(\R^{n \times m}\) & The space of real-valued matrices with \(n\) rows and \(m\) columns. \\
   \(\Id{n}\) & The \(n \times n\) identity matrix. \\
   \(\Zmat{n \times m}\) & The \(n \times m\) matrix of all zeros. \\
   \(M_i\) & If \(M\) is a vector, the \(i\)th element of \(M\). \\
         & If \(M\) is a matrix, the \(i\)th column of \(M\). \\
   \(M_{i,j}\) & The value of row \(i\), column \(j\) for the matrix \(M\). \\
   \(\dot{x}\) & Derivative of \(x\) with respect to time \(t\). \\
   \(\nabla_v F\) & If \(F\) is \(\R\)-valued, the gradient of \(F\) with respect to \(v\). \\
            & If \(F : \R^m \rightarrow \R^{n\times n}\), the block matrix gradient \((\pdiff{F}{v_1},\ldots,\pdiff{F}{v_m}) \in \R^{nm \times n}\). \\
   \(dF_v\) & Total differential (Jacobian) of \(F\), equivalent to \((\nabla_v F)\tpose\). \\
   \(\Hess F\) & If \(F : \R^n \rightarrow \R\), the \(n \times n\) Hessian matrix of double derivatives of \(F\). \\
               & If \(F : \R^n \rightarrow \R^k\), the block matrix \((\Hess F_1, \ldots, \Hess F_k) \in \R^{n \times nk}\). \\
   \(\partial_v \partial_w F\) & Derivative matrix of \(F : \R^n \times \R^m \rightarrow \R\), with \((i,j)\) element \(\ppdiff{F}{v_i}{w_j}\). \\
   \(\delta_{i,j}\) & The Kronecker delta: 1 if \(i = j\) and 0 otherwise. \\
   \(\otimes\) & The matrix kronecker product (see Appendix \ref{appendix:matrix-kron-prod}). \\
   \bottomrule
\end{tabular}

### 2020-07-30 latex/vnhcs.tex
Major edits to VNHC
Added several new VNHC-related references, updated the whole chapter of
VNHCs, and implemented my new notation using the matrix kronecker
product. Also added further justfication as to why we have our own
definition of VNHCs, which needs more work. Next, we need to update the
proof on why a VNHC is regular iff it is h(q,pu) and satisfies that
matrix rank definition.

\section{Preliminaries on Analytical Mechanics}
A mechanical system can be represented by \(N\) point masses, where each point
represents the center of mass of a physical body, along with \(r\)
\textit{equations of constraint} (EOC) which model the physical restrictions
between these masses.
The position of each point mass is described using three cartesian coordinates (one
for each spatial axis), so the system as a whole can be described by a vector in
\(\R^{3N}\) with \(r\) EOC. 
The dynamics of the system are computed by deriving the \(3N\)
\textit{equations of motion} (EOM) produced by Newton's second law \(F = m a\).
While this technique works for simple systems, it is impossible to
apply to a majority of mechanical systems since most of the forces 
on the system are not explicitly known.

@@

For the mechanical systems of interest in this thesis, we assume that
each generalized coordinate \(q_i\) represents either the distance or the angle
between two parts of the system.
Mathematically, each \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to collect the coordinates into a \textit{configuration} 
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) 
where the \textit{configuration manifold} \(\mathcal{Q}\) of the system is a
so-called \textit{generalized cylinder}:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity} of the system. For arbitrary systems,
the space of allowable velocities depends on the current configuration of the
system.  However, since \(\mathcal{Q}\) is a generalized cylinder,
\(\dot{q} \in \R^n\).
The combined vector \((q,\dot{q}) \in \mathcal{Q}\times\R^n\) is called a 
\textit{state} of the system.

@@

% TODO: Replace [p D_i p] here with our new notation
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p} = -\frac{1}{2} \left[p\tpose \pdiff{D^{-1}}{q_i}(q) p\right] 
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

A set of coordinates \(\{q,p\}\) which satisfy Hamilton's equations 
under the Hamiltonian \(\mathcal{H}(q,p)\) are
said to be \textit{canonical coordinates} for the system. A change of
coordinates \((q,p) \rightarrow (Q,P)\) is a \textit{canonical
transformation} if \((Q,P)\) are canonical coordinates with 
\(H(Q,P) = \mathcal{H}\left(q(Q,P), p(Q,P)\right)\).

@@

\begin{defn}
    A set of canonical coordinates \(\{q,p\}\) for a Hamiltonian system 
    \(\mathcal{H}\) are said to be \textit{simply actuated coordinates} if the
    input matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \simpleB    
    \]
\end{defn}
\begin{defn}
    A Hamiltonian system is said to be \textit{simply actuated} if there exists
    a canonical transformation from any set of canonical coordinates 
    \(\{q,p\}\) into simply actuated coordinates for \(\mathcal{H}\).
\end{defn}

@@

Note that if \(k = (n-1)\), the existence of a left annihilator \(A^0\) for
\(B\) implies \(B^\perp := A^0/\norm{A^0}\) satisfies Assumption
\ref{assm:B-perp}.

\begin{assm}\label{assm:B-orthogonal}
    Assume without loss of generality that the input matrix \(B\) is left
    semi-orthogonal.
    That is, assume \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{k \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 & 0 \\
        0 & \sigma_2 & \cdots & 0 & 0 \\
        \vdots & & \ddots &  & \vdots \\
        0 & 0 & \cdots & 0 & \sigma_k \\
          & - & \Zmat{(n-k)\times k} & -  &\\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\) because \(B\) is full-rank.
\textbf{SOURCE}
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau} \in \R^k\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

@@

\begin{defn}
    Let \(\{q,p\}\) be canonical coordinates for a Hamiltonian mechanical system
    and define the coordinate transformation 
    \((q_u,q_a) = \mathbf{B}q\), where \(q_u = B^\perp q\)
    and \(q_a = B\tpose q\).
    We call \(q_u\) the \textit{unactuated coordinates} and \(q_a\) the
    \textit{actuated coordinates} of the system. 
    Their corresponding conjugate of momenta are 
    \((p_u,p_a) = \mathbf{B}p\).
\end{defn}

The following theorem shows that these new coordinates are simply actuated
coordinates, so that \(\tau_i\) only affects \(p_{a,i}\)
and all \(p_u\) are unaffected by the input forces.

% TODO: Replace the [P M_i P] with new notation
\begin{thm}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the change of coordinates 
    \(\{Q = (q_u, q_a), P = (p_u,p_a)\}\) 
    is a canonical transformation of the
    original Hamiltonian system (\ref{eqn:hamiltonian}) into
    the simply actuated Hamiltonian system (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2}
                \left[P\tpose \pdiff{\Minv}{Q_i}(Q) P \right] 
                - \nabla_Q V(Q) + 
                \begin{bmatrix}
                    \Zmat{(n-k)\times k} \\
                    \Id{k} \\
                \end{bmatrix} \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    The Poisson bracket between \(f(q,p)\) and \(g(q,p)\) is defined by
    \cite{landau_mechanics} as follows:
    \[
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
                \pdiff{f}{q_i}\pdiff{g}{p_i}
    \]
    For any constant matrix \(A\), the transformation
    \(\{Q = Aq, P = Ap\}\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, for our new coordinates \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\),
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(\mathbf{B}\) is orthogonal,
    \((\mathbf{B}_i)\tpose \mathbf{B}\tpose_j = (\mathbf{B}_i)\tpose (\mathbf{B}^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P\) and \(Q\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}\tpose_{m,j} \\
                  &= (\mathbf{B}_i)\tpose \mathbf{B}\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

@@

    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the coordinates \(\{q_u,q_a,p_u,p_a\}\) are simply actuated coordinates for 
    \(\hat{H}\) as desired.
\end{proof}

\section{Virtual Nonholonomic Constraints}
% TODO: Motivation for VNHCs, they are an extension of VHCs, etc.
Let us imagine a child on a swing, who wants to go as high as possible. The
child will push off the ground and start a swing with a very small oscillation,
then extend and retract their feet appropriately to gain energy. 
If a roboticist were designing a machine to do this task, they might design a
control mechanism which makes the legs track a trajectory over time. For
example, they might tell the robot to extend and retract its legs every two
seconds. In ideal situtations, this technique would work perfectly because the
leg motion is synchronized with the swing to gain energy as fast as possible.

Unfortunately for engineers, reality does not provide ideal scenarios very often. 
Most children have an adult pushing the swing to help them go higher; or perhaps
the child is swinging on a windy day. In either case, they adjust their leg
motion accordingly and do not keep track of time when kicking their legs. This
standard control technique of tracking a function of time, known as
\textit{trajectory tracking}, would not work because the disturbance affecting
the swing will desynchronize the motion of the legs with the swing and stop the
energy-gaining effects.

This is a common issue in the control of many biologically-inspired systems. 
One method which can provide more realistic, robust motion is the method of
of virtual holonomic constraints (VHCs). Instead of a robot's actuators tracking
a trajectory over time, VHCs use the actuators to enforce a relation
\(h(q) = 0\) of the configuration \cite{vhcs_for_el_systems}. 
This method has provided incredible results in the development of 
walking robots \cite{vhc_robotic_walking, vhc_stable_walking}, 
vehicle motion \cite{vhc_bicycle, vhc_helicopter}, 
and has even been used to design a snake-like robot that can swim
\cite{vhc_snake}.

Recent papers have attempted to extend these results to use the robot's
actuators to enforce a relation 
\(h(q,\dot{q}) = 0\) of the full state. Since these relations use both
the configuration and generalized velocity, they are called
\textit{nonholonomic} constraints. Furthermore, they are called \textit{virtual}
because the constraint is enforced by the robot's actuators.
We will give our own characterization of these virtual nonholonomic constraints
in the Hamiltonian framework, using the phase \((q,p)\) of a Hamiltonian
system instead of the state \((q,\dot{q})\) of a lagrangian system.

@@

Defining the error term  \(e = h(q,p)\), stabilizing \(\Gamma\) is equivalent to 
solving for \(\tau(q,p)\) which drives \(e \rightarrow 0\)
and \(\dot{e} \rightarrow 0\).
Let us imagine that we have no further structure on the VNHC. 
Then \(\dot{e} = dh_q \dot{q} + dh_p \dot{p}\), so \(\tau\) appears inside
\(\dot{e}\). To solve for \(\tau\) explicitly, we must have that \(dh_p\) is
invertible, which means that \(h(q,p)\) is strictly monotonic in \(p\).

This is a very restrictive condition, which many VNHCs will not satisfy. For
this reason, it is preferable to have the torque \(\tau\) appear after two 
derivatives of \(e\) so that there is more freedom in the types of constraints
one can use.
The terminology for this is to say that \(e\) is of \textit{relative degree}
\(\{2,2,\ldots,2\}\). We thus define a special type of VNHC which satisfies this
property.

@@

\citeauthor{hybrid_zero_dynamics_bipedal_nhvcs} showed that relations which use only
the unactuated momentum \(p_u\) result in \(\tau\) appearing only after two
derivatives of \(e\) \cite{hybrid_zero_dynamics_bipedal_nhvcs}. 
In the remainder of this chapter, we will formalize and generalize this idea 
under the following assumption.

% TODO: Write the [p' M p] matrix in our nice nottation
\begin{assm}\label{assm:H-is-simply-actuated}
    The mechanical system under consideration is an underactuated 
    Hamiltonian system of order \((n-k) \geq 0\) in simply
    actuated coordinates \(\{q = (q_u,q_a), p = (p_u, p_a)\}\):
    \begin{align*}
        \mathcal{H}(q,p) &= p\tpose \Minv(q) p + V(q) \\
         &\begin{cases}
            \dot{q} = \Minv p \\
            \dot{p} = -\frac{1}{2} ? - \nabla_q V(q) + \simpleB \tau \\
        \end{cases}
    \end{align*}
\end{assm}

% TODO: write out ??? = [p' dM/(dqu_i)] the matrix of size nxn
\begin{prop}
    A VNHC of order \(k\) is regular if and only if \(dh_{p_a} = 0\) 
    (\ie \(h(q,p) \equiv h(q,p_u)\)) and
    \[
        \rank{\left(dh_q \Minv(q) + dh_{p_u} ??? \right)\simpleB} = k
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{prop}
\begin{proof}
\end{proof}

>>>>

\section{Preliminaries on Analytical Mechanics}
A mechanical system can be represented by \(N\) point masses where each point
represents the center of mass of a physical body, along with \(r\)
\textit{equations of constraint} (EOC) which model the physical restrictions
between these masses.
The position of each point mass is described using three cartesian coordinates (one
for each spatial axis), so the system as a whole can be described by a vector in
\(\R^{3N}\) with \(r\) EOC. 
The dynamics of the system are computed by deriving the \(3N\)
\textit{equations of motion} (EOM) produced by Newton's second law \(F = m a\).
While this technique works for simple systems, it is impossible to
apply to a majority of mechanical systems since most of the forces 
on the system are not explicitly known.

@@

For the robotic systems of interest in this thesis, we assume that
each generalized coordinate \(q_i\) represents either the distance or the angle
between two parts of the system.
Mathematically, each \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to collect the coordinates into a \textit{configuration} 
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) 
where the \textit{configuration manifold} \(\mathcal{Q}\) of the system is a
so-called \textit{generalized cylinder}:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity} of the system. For arbitrary systems,
the space of allowable velocities depends on the current configuration of the
system.  However, since \(\mathcal{Q}\) is a generalized cylinder, we find that 
\(\dot{q} \in \R^n\).
The combined vector \((q,\dot{q}) \in \mathcal{Q}\times\R^n\) is called a 
\textit{state} of the system.

@@

% TODO: Replace [p D_i p] here with our new notation
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D\inv(q)p \\
        \dot{p} = -\frac{1}{2} (I_n \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

A set of coordinates \(\{q,p\}\) which satisfy Hamilton's equations 
under the Hamiltonian \(\mathcal{H}\) are
said to be \textit{canonical coordinates} for the system. A change of
coordinates \((q,p) \rightarrow (Q,P)\) is a \textit{canonical
transformation} if \(\{Q,P\}\) are canonical coordinates with Hamiltonian
\(H(Q,P) = \mathcal{H}\left(q(Q,P), p(Q,P)\right)\).

@@

\begin{defn}
    Let \(\mathcal{H}\) be an underactuated Hamiltonian system 
    \(\mathcal{H}\) with degree of underactuation \((n-k)\geq 0\). 
    A pair of canonical coordinates \(\{q,p\}\) for this system
    are said to be \textit{simply actuated coordinates} if the
    input matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \simpleB    
    \]
    The first \((n-k)\) coordinates \(q_u\) are called the \textit{unactuated
    coordinates}, while the remaining \(k\) coordinates \(q_a\) are called the
    \textit{actuated coordinates}. We write \(q = (q_u, q_a)\) and the
    corresponding conjugate of momenta \(p = (p_u, p_a)\) are the called the
    unactuated and actuated momenta (respectively).
\end{defn}
\begin{defn}
    A Hamiltonian system is said to be \textit{simply actuated} if there exists
    a canonical transformation from any set of canonical coordinates 
    \(\{q,p\}\) into simply actuated coordinates for \(\mathcal{H}\).
\end{defn}

@@

Note that if \(k = (n-1)\), the existence of any left annihilator \(A^0\) for
\(B\) implies the left annihilator \(B^\perp := A^0/\norm{A^0}\) satisfies Assumption
\ref{assm:B-perp}.

\begin{assm}\label{assm:B-orthogonal}
    Assume without loss of generality that the input matrix \(B\) is left
    semi-orthogonal.
    That is, assume \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{k \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 \\
        0 & \sigma_2 & \cdots & 0 \\
        \vdots & & \ddots & \vdots \\
        0 & 0 & \cdots & \sigma_k \\
        - &   & \Zmat{(n-k)\times k} & -  \\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\) because \(B\) is full-rank \textbf{SOURCE?}.
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau} \in \R^k\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

@@

The following theorem shows that \(\mathbf{B}\) provides a canonical
transformation into simply actuated coordinates, so that all unactuated momenta
are unaffected by the input forces.

% TODO: Replace the [P M_i P] with new notation
\begin{thm}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system (\ref{eqn:hamiltonian}) is
    simply actuated with simply actuated coordinates 
    \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\). The resulting dynamics are 
    given by (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2} (I_n \otimes P\tpose) \nabla_Q \Minv(Q) P
                - \nabla_Q V(Q) + \simpleB \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &:= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &:= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    The Poisson bracket between the functions \(f(q,p)\) and \(g(q,p)\) is defined by
    \cite{landau_mechanics} as follows:
    \[
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
                \pdiff{f}{q_i}\pdiff{g}{p_i}
    \]
    For any constant matrix \(A\), the transformation
    \(\{Q = Aq, P = Ap\}\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, for our new coordinates \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\),
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(\mathbf{B}\) is orthogonal,
    \((\mathbf{B}_i)\tpose \mathbf{B}\tpose_j = (\mathbf{B}_i)\tpose (\mathbf{B}^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P\) and \(Q\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}\tpose_{m,j} \\
                  &= (\mathbf{B}_i)\tpose \mathbf{B}\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

@@

    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the coordinates \(\{Q = (q_u,q_a), P = (p_u,p_a)\}\) are simply actuated coordinates for 
    \(\hat{H}\) as desired.
\end{proof}

\section{Virtual Nonholonomic Constraints}
% TODO: Motivation for VNHCs, they are an extension of VHCs, etc.
Let us imagine a child on a swing, who wants to go as high as possible. The
child will push off the ground and start swinging with small oscillations,
then extend and retract their feet appropriately to gain energy. 
If a roboticist were designing a machine to do this task, they might design a
control mechanism which makes the legs track a trajectory over time. For
example, they might tell the robot to extend and retract its legs every two
seconds. In ideal situtations, this technique would work perfectly because the
leg motion is synchronized with the swing to gain energy as fast as possible.

Most children have an adult pushing the swing to help them go higher; or perhaps
the child is swinging on a windy day. In either case, they adjust their leg
motion accordingly and do not keep track of time when kicking their legs. Te
standard control technique of tracking a function of time, known as
\textit{trajectory tracking}, would not work because the disturbance affecting
the swing will desynchronize the motion of the legs with the swing and stop the
energy-gaining effects.

This is a common issue in the control of many biologically-inspired systems. 
One method which can provide more realistic, robust motion is the method of
of virtual holonomic constraints (VHCs). Instead of a robot's actuators tracking
a trajectory over time, VHCs use the actuators to enforce a relation
\(h(q) = 0\) of the configuration \cite{vhcs_for_el_systems}. 
This method has provided incredible results in the development of 
walking robots \cite{vhc_robotic_walking, vhc_stable_walking}, 
vehicle motion \cite{vhc_bicycle, vhc_helicopter}, 
and has even been used to design a snake-like swimming robot
\cite{vhc_snake}.

Many authors have attempted to extend these results to enforce a relation 
\(h(q,\dot{q}) = 0\) of the full state. Since these relations use actuators to
restrict both the configuration and generalized velocity, they are called
virtual \textit{nonholonomic} constraints. This idea has been used for
human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and improving bipedal locomotion \cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking,output_nhvc_bipedal_control}.

Unlike the theory of VHCs, there does not appear to be a standard definition of virtual
nonholonomic constraints. One issue is that Lagrangian dynamics is not
particularly suited to finding controllers which stabilize desired constraints. Because
of this, all the applications listed above use their own definition of a virtual
nonholonomic constraint, which makes it difficult to compare and analyze their
work.

This section will provide a new characterization of virtual nonholonomic
constraints. The goal is to provide a consistent, rigorous foundation for
designing constraints on a general class of systems.

@@

If we define the error term  \(e = h(q,p)\), stabilizing \(\Gamma\) is
equivalent to solving for \(\tau(q,p)\) which drives \(e \rightarrow 0\) and
\(\dot{e} \rightarrow 0\).
Let us imagine that we have no further structure on the VNHC. 
Then \(\dot{e} = dh_q \dot{q} + dh_p \dot{p}\), so \(\tau\) appears inside
\(\dot{e}\). To solve for \(\tau\) explicitly, we must have that \(dh_p\) is
invertible for all \(p\), which is the same as requiring that \(h(q,p)\) is
strictly monotonic in \(p\).

This is a very restrictive condition, which many VNHCs will not satisfy. For
this reason, it is preferable to have the torque \(\tau\) appear after two 
derivatives of \(e\) so that there is more freedom in the types of constraints
one can use.
Mathematically, if \(\tau\) appears only after two derivatives, one says 
that \(e\) is of \textit{relative degree} \(\{2,2,\ldots,2\}\). 
We thus define a special type of VNHC which satisfies this
property.

@@

Modifying the results of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
into the Hamiltonian framework, one observes that
relations which use only the unactuated momentum \(p_u\) result in \(\tau\)
appearing only after two derivatives of \(e\).
To be able to use \(p_u\), we will continue with the following assumption for
the rest of the chapter.

% TODO: Write the [p' M p] matrix in our nice nottation
\begin{assm}\label{assm:H-is-simply-actuated}
    The mechanical system under consideration is an underactuated 
    Hamiltonian system of order \((n-k) \geq 0\) in simply
    actuated coordinates \(\{q = (q_u,q_a), p = (p_u, p_a)\}\):
    \begin{align*}
        \mathcal{H}(q,p) &= p\tpose \Minv(q) p + V(q) \\
         &\begin{cases}
            \dot{q} = \Minv p \\
            \dot{p} = -\frac{1}{2} (I_n \otimes p\tpose) \nabla_q \Minv(q) p
            - \nabla_q V(q) + \simpleB \tau \\
        \end{cases}
    \end{align*}
\end{assm}

% TODO: write out ??? = [p' dM/(dqu_i)] the matrix of size nxn
\begin{prop}
    A VNHC of order \(k\) is regular if and only if \(dh_{p_a} = 0\) 
    and
    \[
        \rank{\left(dh_q \Minv(q) + 
          dh_{p_u} (I_n \otimes p\tpose)\nabla_{q_u}\Minv(q) 
         \right)\simpleB} = k
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{prop}
\begin{proof}
\end{proof}
